% ─────────────────────────────────────────────────────────────────────────────
% Ternlang: A Full-Stack Balanced Ternary Execution Architecture
% for Sparse Neural Inference and Ambiguity-Aware Agent Systems
%
% RFI-IRFOS Ternary Intelligence Stack (TIS)
% Author: Simeon Kepp, RFI-IRFOS
%
% Submission target: arXiv cs.PL / cs.AR / cs.NE
% Format: IEEE two-column (IEEEtran) or ACM (acmart)
% ─────────────────────────────────────────────────────────────────────────────

\documentclass[journal,10pt,twocolumn]{IEEEtran}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{multirow}
\usepackage{array}

% ── Ternlang code style ───────────────────────────────────────────────────────
\definecolor{ternkw}{RGB}{0,164,140}
\definecolor{terncmt}{RGB}{120,120,150}
\definecolor{ternlit}{RGB}{220,80,80}
\definecolor{ternbg}{RGB}{18,18,30}
\definecolor{ternfg}{RGB}{220,220,230}

\lstdefinelanguage{ternlang}{
  morekeywords={fn,let,match,if,for,in,loop,break,return,
                agent,spawn,send,await,struct,trit,trittensor,agentref,
                truth,hold,conflict,consensus,invert,matmul,sparsity,cast},
  morecomment=[l]{//},
  morestring=[b]",
  sensitive=true
}

\lstset{
  language=ternlang,
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{ternkw}\bfseries,
  commentstyle=\color{terncmt}\itshape,
  stringstyle=\color{ternlit},
  showstringspaces=false,
  breaklines=true,
  frame=single,
  numbers=left,
  numberstyle=\tiny\color{terncmt},
  numbersep=4pt,
  xleftmargin=10pt
}

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}

% ─────────────────────────────────────────────────────────────────────────────

\title{Ternlang: A Full-Stack Balanced Ternary Execution Architecture
       for Sparse Neural Inference and Ambiguity-Aware Agent Systems}

\author{Simeon~Kepp,~\IEEEmembership{RFI-IRFOS}%
\thanks{Correspondence: RFI-IRFOS Ternary Intelligence Stack.
Repository: \url{https://github.com/eriirfos-eng/ternary-intelligence-stack--tis-}}%
}

\begin{document}

\maketitle

% ─────────────────────────────────────────────────────────────────────────────
\begin{abstract}
We present \textbf{Ternlang}, the first complete software stack for
balanced ternary computing: a domain-specific language, bytecode compiler,
stack-based virtual machine (BET VM), hardware description language backend,
distributed actor runtime, and machine learning inference kernels—all unified
under a single coherent architecture. The foundational primitive is the
\emph{trit} $t \in \{-1, 0, +1\}$, where the value $0$ represents an
\emph{active neutral state} rather than absence, enabling three-valued logic
that is structurally superior to binary for ambiguity-aware reasoning.

The principal contribution is \textbf{TSPARSE\_MATMUL}: a first-class VM
opcode that elides multiply operations against zero-weighted (``hold'') trit
elements, surfacing at the instruction-set level a property that
BitNet-style~\cite{ma2024era} ternary quantization reveals in neural weights.
Empirical evaluation on quantized weight matrices of $512 \times 512$
elements yields $56.2\%$ sparsity and a $2.27\times$ reduction in multiply
operations versus dense execution; a Compressed Sparse Column kernel with
Rayon parallelism achieves up to $122\times$ speedup at $99\%$ sparsity—
without approximate arithmetic or hardware reconfiguration.

We further define the \textbf{Balanced Ternary Execution (BET) ISA}, a
formal 2-bit-encoded instruction set with 51 opcodes spanning arithmetic,
tensor operations, actor messaging, and control flow; synthesise it to
Verilog-2001 with per-cell clock-gating on zero-weight elements; and
demonstrate interoperability with existing ternary computing efforts via the
\texttt{ternlang-compat} bridge crate.

Beyond the execution substrate, we introduce two higher-level contributions.
The \textbf{Ternary AI Reasoning Toolkit} (Phase~8) provides five structural
primitives—iterative deliberation, coalition voting, multi-dimensional action
gating, scalar-to-temperature bridging, and hallucination scoring—that make
any AI agent's decision loop architecturally ternary rather than merely
prompted. The \textbf{MoE-13 Ternary Orchestrator} (Phase~9) implements a
Mixture-of-Experts head with dual-key synergistic routing, $1{+}1{=}3$
emergent triad synthesis, a hard-wired safety veto, and a three-tier memory
mesh, realising the MoE-13 architecture~\cite{moe13_2026} in 146 passing
tests across the full workspace. All nine crates are published to
\texttt{crates.io} under open-core licensing. A companion corpus of
twenty executable \texttt{.tern} programs demonstrates the hold state
across domains spanning aerospace, medicine, distributed systems, civic
governance, and AI agent design.
\end{abstract}

\begin{IEEEkeywords}
balanced ternary, ternary logic, sparse inference, BitNet, domain-specific
language, virtual machine, actor model, FPGA synthesis, Verilog,
mixture-of-experts, AI reasoning, ternary orchestration
\end{IEEEkeywords}

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:intro}

The computational substrate underlying modern artificial intelligence is
binary. Floating-point arithmetic, two-state memory, and Boolean logic gates
have driven five decades of progress—but they introduce a fundamental
representational mismatch when modelling systems that are inherently
three-valued: \emph{affirmed}, \emph{denied}, and \emph{undecided}.

Clinical diagnosis, legal reasoning, sensor fusion under noise, and
multi-agent consensus all require a native neutral state that binary computing
forces to encode as a special case—null pointers, NaN, sentinel values, or
probabilistic scores collapsed to a threshold. Each encoding is a workaround
for an absent primitive.

Balanced ternary~\cite{knuth1997art} provides that primitive. A \emph{trit}
$t \in \{-1, 0, +1\}$ carries three symmetric values. The neutral value $0$
is \emph{active}—a deliberate state of hold, not an empty bit pattern.
Balanced ternary arithmetic is self-complementing: negation requires no
special-case handling. And at the scale of modern neural networks, where
BitNet~\cite{ma2024era} and related work show that ternary-quantized weights
preserve accuracy with dramatically reduced computation, the case for a
ternary-native execution substrate is both theoretical and empirical.

Despite this, the ternary computing field remains fragmented: hobbyist
emulators, academic EDA tools for memristor hardware~\cite{bos2020ternary},
isolated Lisp interpreters, and hardware simulators without compiler support.
No project provides the full vertical stack.

\textbf{Ternlang} fills this gap. Our contributions are:

\begin{enumerate}
  \item A \emph{language design} for balanced ternary: three-way exhaustive
        pattern matching, first-class trit tensors, and an actor model for
        ternary message passing (\S\ref{sec:language}).
  \item The \textbf{BET ISA}: a formal 2-bit-encoded instruction set with
        opcode coverage from arithmetic through tensor operations and
        distributed agent control (\S\ref{sec:isa}).
  \item \textbf{TSPARSE\_MATMUL}: a VM opcode that skips zero-weight
        multiplications at the instruction level, realising the sparsity
        benefit of ternary quantization without software overhead
        (\S\ref{sec:sparse}).
  \item A \textbf{Verilog-2001 hardware backend} with synthesisable sparse
        matmul array and full BET processor, plus a cycle-accurate RTL
        simulator in pure Rust (\S\ref{sec:hdl}).
  \item A \textbf{Ternary AI Reasoning Toolkit}: five structural
        primitives—deliberation, coalition vote, action gate, scalar
        temperature, and hallucination score—that make any AI agent's
        decision loop architecturally ternary (\S\ref{sec:reasoning}).
  \item The \textbf{MoE-13 Ternary Orchestrator}: dual-key synergistic
        routing, $1{+}1{=}3$ emergent triad synthesis, a hard safety veto,
        and three-tier memory mesh, exposed as MCP tool calls
        (\S\ref{sec:moe}).
  \item \textbf{Ecosystem bridges} connecting existing ternary projects to
        the BET VM as a common runtime (\S\ref{sec:ecosystem}).
  \item A \textbf{reference example corpus}: twenty executable
        \texttt{.tern} programs demonstrating the hold state in aerospace,
        medicine, distributed systems, hardware pipelines, civic governance,
        finance, and AI agent design—published open-source alongside all
        nine workspace crates on \texttt{crates.io}.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Background: Balanced Ternary}
\label{sec:background}

\subsection{Trit arithmetic}

A \emph{trit} $t \in T = \{-1, 0, +1\}$ participates in the following
complete operations~\cite{knuth1997art}:

\begin{definition}[Ternary addition with carry]
For trits $a, b \in T$, define $s = \mathrm{sum}(a,b)$ and
$c = \mathrm{carry}(a,b)$ such that $a + b = 3c + s$, where
$s, c \in T$.
\end{definition}

The complete sum/carry table is:

\begin{table}[h]
\centering
\small
\caption{Balanced ternary addition (sum, carry)}
\label{tab:add}
\begin{tabular}{c|ccc}
\toprule
$a \backslash b$ & $-1$ & $0$ & $+1$ \\
\midrule
$-1$ & $(+1,-1)$ & $(-1,\ 0)$ & $(0,\ 0)$ \\
$\ 0$ & $(-1,\ 0)$ & $(0,\ 0)$ & $(+1,\ 0)$ \\
$+1$ & $(0,\ 0)$ & $(+1,\ 0)$ & $(-1,+1)$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{definition}[Ternary multiplication]
$\mathrm{mul}(a,b) = a \cdot b$ under integer multiplication,
restricted to $T$: $\mathrm{mul}(a,b) \in T$ since $|a|,|b| \leq 1$.
\end{definition}

\begin{definition}[Consensus (ternary OR)]
$\mathrm{cons}(a,b) = a$ if $a = b$, else $0$.
\end{definition}

\begin{definition}[Negation]
$\mathrm{neg}(t) = -t$. Note $\mathrm{neg}(0) = 0$.
\end{definition}

\subsection{The 2-bit BET encoding}

Hardware naturally operates in binary. We encode trits as 2-bit pairs:

\begin{table}[h]
\centering
\small
\caption{BET 2-bit trit encoding}
\label{tab:encoding}
\begin{tabular}{ccc}
\toprule
Bits & Value & Meaning \\
\midrule
\texttt{0b01} & $-1$ & conflict \\
\texttt{0b10} & $+1$ & truth \\
\texttt{0b11} & $\ 0$ & hold \\
\texttt{0b00} & --- & FAULT (invalid) \\
\bottomrule
\end{tabular}
\end{table}

This encoding is \emph{not} two's complement. The bit pattern \texttt{0b11}
for zero is chosen so that the all-ones reset state of a register file
initialises every register to hold—the semantically correct neutral value—
without special reset logic. Negation maps to a simple bit swap:
\texttt{0b01} $\leftrightarrow$ \texttt{0b10}, leaving \texttt{0b11}
invariant.

\begin{proposition}[Negation as bit swap]
For encoding $e(t)$ as above, $e(\mathrm{neg}(t)) = \{e(t)[0], e(t)[1]\}$
(swap bits).
\end{proposition}

% ─────────────────────────────────────────────────────────────────────────────
\section{The Ternlang Language}
\label{sec:language}

\subsection{Design principles}

Ternlang is statically typed, expression-oriented, and compiled to BET
bytecode. Three design decisions distinguish it from binary-native languages
adapted to ternary:

\textbf{Exhaustive three-way matching.} Every \texttt{match} expression must
cover all three trit arms. The compiler rejects non-exhaustive matches at
parse time, eliminating an entire class of runtime error present in languages
that bolt ternary logic onto binary pattern matching.

\textbf{$0$ is active.} The type system and semantics assign distinct meaning
to $-1$ (conflict), $0$ (hold, active neutral), and $+1$ (truth). There is no
null, no undefined, no NaN. A trit always carries a definite value.

\textbf{Sparsity is a language feature.} The \texttt{@sparseskip} directive
marks a tensor operation as sparse-aware, routing the compiler to emit
\texttt{TSPARSE\_MATMUL} instead of \texttt{TMATMUL}. Sparsity is expressed
in the source language, not discovered by the optimizer.

\subsection{Core language constructs}

\begin{lstlisting}[caption={Ternary classifier with exhaustive match}]
fn classify(signal: trit) -> trit {
    match signal {
        -1 => conflict()  // disagreement confirmed
         0 => hold()      // awaiting evidence
        +1 => truth()     // assertion confirmed
    }
}
\end{lstlisting}

\begin{lstlisting}[caption={Sparse matrix multiply with @sparseskip}]
// Route to TSPARSE_MATMUL at the ISA level
@sparseskip
let output: trittensor<8 x 8> =
    matmul(input, weights);
\end{lstlisting}

\begin{lstlisting}[caption={Actor model for ternary message passing}]
agent Voter {
    fn handle(msg: trit) -> trit {
        consensus(msg, hold())
    }
}

let v: agentref = spawn Voter;
send v truth();
let decision: trit = await v;
\end{lstlisting}

\subsection{Type system}

The core types are \texttt{trit} (a single balanced ternary value),
\texttt{trittensor<N x M>} (an $N \times M$ matrix of trits stored on the
tensor heap), and \texttt{agentref} (a handle to a running actor instance,
local or remote). Struct types with trit/tensor fields are supported via
field-name mangling in the register allocator.

% ─────────────────────────────────────────────────────────────────────────────
\section{The BET Instruction Set Architecture}
\label{sec:isa}

\subsection{Machine model}

The BET VM is a stack-based machine with:

\begin{itemize}
  \item \textbf{27 registers} (2 bits each), reset to \texttt{0b11} (hold)
  \item \textbf{Value stack}: unbounded, stores \texttt{Value} tagged union
  \item \textbf{Tensor heap}: indexed array of $N \times M$ trit matrices
  \item \textbf{Call stack}: return address stack for \texttt{TCALL}/\texttt{TRET}
  \item \textbf{Agent table}: maps type IDs to handler addresses and mailboxes
  \item \textbf{Carry register}: overflow from \texttt{TADD}
\end{itemize}

The choice of 27 registers reflects the ternary motif: $27 = 3^3$.

\subsection{Instruction encoding}

Instructions are variable-length, with a 1-byte opcode followed by 0–2
operand bytes. Table~\ref{tab:isa} lists the complete opcode map.

\begin{table}[h]
\centering
\small
\caption{BET ISA opcode reference (selected)}
\label{tab:isa}
\begin{tabular}{lll}
\toprule
Opcode & Mnemonic & Stack effect \\
\midrule
\texttt{0x00} & THALT            & --- \\
\texttt{0x01} & TPUSH \textit{t} & $\to t$ \\
\texttt{0x02} & TADD             & $a\ b \to s\ c$ \\
\texttt{0x03} & TMUL             & $a\ b \to t$ \\
\texttt{0x04} & TNEG             & $t \to \mathrm{neg}(t)$ \\
\texttt{0x05} & TJMP\_POS \textit{addr} & $t \to$ (jmp if $t=+1$) \\
\texttt{0x06} & TJMP\_ZERO \textit{addr}& $t \to$ (jmp if $t=0$) \\
\texttt{0x07} & TJMP\_NEG \textit{addr} & $t \to$ (jmp if $t=-1$) \\
\texttt{0x08} & TSTORE \textit{r}& $t \to$ (reg[$r$]$\leftarrow t$) \\
\texttt{0x09} & TLOAD \textit{r} & $\to$ reg[$r$] \\
\texttt{0x0b} & TJMP \textit{addr}& --- \\
\texttt{0x0e} & TCONS            & $a\ b \to \mathrm{cons}(a,b)$ \\
\texttt{0x0f} & TALLOC \textit{N M} & $\to$ ref \\
\texttt{0x10} & TCALL \textit{addr} & (push PC; jmp) \\
\texttt{0x11} & TRET            & (pop PC; jmp) \\
\texttt{0x20} & TMATMUL         & $r_A\ r_B \to r_C$ \\
\texttt{0x21} & TSPARSE\_MATMUL & $r_A\ r_B \to r_C$ \\
\texttt{0x22} & TIDX            & ref $i\ j \to t$ \\
\texttt{0x23} & TSET            & ref $i\ j\ t \to$ \\
\texttt{0x24} & TSHAPE          & ref $\to N\ M$ \\
\texttt{0x25} & TSPARSITY       & ref $\to$ count \\
\texttt{0x30} & TSPAWN \textit{id} & $\to$ agentref \\
\texttt{0x31} & TSEND           & agentref $m \to$ \\
\texttt{0x32} & TAWAIT          & agentref $\to t$ \\
\bottomrule
\end{tabular}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section{Sparse Ternary Inference}
\label{sec:sparse}

\subsection{Ternary quantization}

BitNet-style ternary weight quantization~\cite{ma2024era} maps floating-point
weights $w \in \mathbb{R}$ to $\hat{w} \in \{-1, 0, +1\}$ using a threshold
$\tau$:

\begin{equation}
  \hat{w} = \begin{cases}
    +1 & w >  \tau \\
     0 & |w| \leq \tau \\
    -1 & w < -\tau
  \end{cases}
  \qquad \tau = \frac{1}{2} \cdot \mathbb{E}[|w|]
  \label{eq:quantize}
\end{equation}

The resulting weight distribution is heavily concentrated at $0$ (hold):
typical language model weights at BitNet scale show $55$–$65\%$ zero elements
after quantization with $\tau$ as defined in~(\ref{eq:quantize}).

\subsection{The @sparseskip directive and TSPARSE\_MATMUL}

The key insight is that a multiply against a zero-weight trit produces zero
regardless of the input value:

\begin{equation}
  \mathrm{mul}(a, 0) = 0 \quad \forall a \in T
  \label{eq:zero-mul}
\end{equation}

In a dense $N \times M$ matrix multiply, every element contributes
$N \cdot M$ multiplications. After ternary quantization with sparsity $\rho$
(fraction of zero-weight elements), only $(1-\rho) \cdot N \cdot M$
multiplications have non-trivial results.

\texttt{TSPARSE\_MATMUL} (opcode \texttt{0x21}) implements a sparse
inner-product loop that tests the weight trit before executing the multiply:

\begin{lstlisting}[language=C, caption={TSPARSE\_MATMUL pseudocode}]
for i in 0..N:
  for j in 0..M:
    for k in 0..K:
      w = W[k][j]
      if w == HOLD: continue   // skip
      acc[i][j] += mul(A[i][k], w)
\end{lstlisting}

The directive \texttt{@sparseskip} in the source language is a compile-time
annotation that causes the codegen to emit \texttt{TSPARSE\_MATMUL} instead of
\texttt{TMATMUL} for the following \texttt{matmul()} call. Sparsity awareness
is thus a \emph{source-language} property surfaced at the ISA level, not a
runtime optimization that may or may not trigger.

\subsection{Benchmark results}

We evaluate sparse versus dense matrix multiply on $512 \times 512$ weight
matrices quantized from normally distributed $\mathcal{N}(0,1)$ float weights
using threshold~(\ref{eq:quantize}).

\begin{table}[h]
\centering
\small
\caption{Sparse vs.\ dense ternary matmul ($512 \times 512$)}
\label{tab:benchmark}
\begin{tabular}{lrr}
\toprule
Metric & Dense & Sparse \\
\midrule
Weight sparsity & 0\% & 56.2\% \\
Multiply operations & 262,144 & 115,343 \\
Skipped operations & 0 & 146,801 \\
Speedup & $1.0\times$ & $\mathbf{2.27\times}$ \\
\bottomrule
\end{tabular}
\end{table}

The $2.27\times$ reduction in multiply operations is achieved without
approximation: the result of \texttt{TSPARSE\_MATMUL} is identical to
\texttt{TMATMUL} by the identity~(\ref{eq:zero-mul}).

% ─────────────────────────────────────────────────────────────────────────────
\section{Hardware Backend}
\label{sec:hdl}

\subsection{Verilog-2001 primitives}

The \texttt{ternlang-hdl} crate generates synthesisable Verilog-2001 modules
from the BET ISA. Each trit becomes a \texttt{[1:0]} bus. The core primitives
are:

\begin{itemize}
  \item \texttt{trit\_neg}: bit swap (\texttt{assign y = \{a[0], a[1]\}})
  \item \texttt{trit\_cons}: \texttt{assign y = (a == b) ? a : 2'b11}
  \item \texttt{trit\_mul}: zero-skip detect with combinational multiply
  \item \texttt{trit\_add}: 9-case Verilog \texttt{case} with carry
  \item \texttt{trit\_reg}: synchronous D-register with reset-to-hold
  \item \texttt{bet\_alu}: top-level ALU selecting among primitives by op code
\end{itemize}

\subsection{Sparse matmul array}

The synthesisable sparse matmul array instantiates an $N \times N$ grid of
processing elements. Each cell contains a weight register and a clock-gate
signal:

\begin{lstlisting}[language=Verilog,
  caption={Per-cell clock gating in sparse matmul}]
wire [1:0] w_ij = weight_reg[i][j];
wire skip = (w_ij == 2'b11);  // hold = zero weight
wire [1:0] contrib = skip ?
    2'b11 :                    // hold if skipped
    trit_mul(a_i, w_ij);
\end{lstlisting}

Clock-gating on the \texttt{skip} signal prevents switching activity in
zero-weight cells, delivering dynamic power reduction proportional to sparsity.

\subsection{BET processor}

The full \texttt{bet\_processor} module wires together:
\begin{itemize}
  \item \texttt{bet\_regfile}: 27-register file ($27 \times 2$ bits)
  \item \texttt{bet\_pc}: 16-bit program counter with load port for jumps
  \item \texttt{bet\_control}: single-cycle decode, all 51 opcodes mapped to
        control signals (alu\_op, reg\_we, mem\_re, mem\_we, pc\_load,
        is\_call, is\_ret, is\_halt, is\_spawn, is\_send, is\_await)
\end{itemize}

An Icarus Verilog simulation wrapper (\texttt{BetSimEmitter}) generates
complete self-contained testbenches from BET bytecode, including inline ROM
initialization, reset sequencing, and VCD waveform export for GTKWave.

% ─────────────────────────────────────────────────────────────────────────────
\section{Actor Model and Distributed Runtime}
\label{sec:actors}

\subsection{Local actors}

Ternlang implements the actor model via three ISA primitives:
\texttt{TSPAWN} creates an agent instance and returns an \texttt{agentref};
\texttt{TSEND} enqueues a trit message in the agent's mailbox; \texttt{TAWAIT}
dequeues a message, invokes the handler, and returns the trit result. The
mailbox is a FIFO (\texttt{VecDeque<Value>}) allocated per agent instance.

\subsection{Distributed actors}

The \texttt{ternlang-runtime} crate extends the actor model to multi-node
systems via TCP transport. A \texttt{TernNode} binds a TCP port, maintains a
peer connection map, and exposes \texttt{remote\_send}/\texttt{remote\_await}
over a newline-delimited JSON wire protocol:

\begin{lstlisting}[language=ternlang,
  caption={Remote actor spawn in ternlang}]
let remote_voter: agentref =
    spawn remote "192.168.1.42:7373" Voter;
send remote_voter truth();
let r: trit = await remote_voter;
\end{lstlisting}

The wire protocol uses four message types (\texttt{Send}, \texttt{Await},
\texttt{Reply}, \texttt{Error}) serialised as tagged JSON objects, enabling
interoperability with non-Rust implementations.

% ─────────────────────────────────────────────────────────────────────────────
\section{Ternary AI Reasoning Toolkit}
\label{sec:reasoning}

Modern AI agents are structurally binary in their decision loops: evidence
is collapsed to a scalar, thresholded, and converted to a Boolean act/wait
flag. The three-valued nature of evidence—affirm, hold, reject—is treated
as a side-effect of confidence scores rather than a first-class type.

Phase~8 of the TIS introduces five primitives in \texttt{ternlang-ml} that
make the decision loop architecturally ternary.

\subsection{Deliberation Engine}

The \texttt{DeliberationEngine} models iterative evidence accumulation using
an exponential moving average (EMA):

\begin{equation}
  S_r = \alpha \cdot e_r + (1-\alpha) \cdot S_{r-1},
  \quad \alpha \in (0, 1]
  \label{eq:ema}
\end{equation}

where $e_r$ is the mean of new evidence signals in round~$r$. The engine
commits to a ternary decision only when the confidence of $S_r$ exceeds a
target threshold; otherwise it remains in the hold state and requests a
further evidence round. This models the human ``let me think about it''
behaviour—a native ternary computation rather than a binary timeout.

\subsection{Coalition Vote}

\texttt{coalition\_vote()} aggregates $N$ independent agent verdicts—each a
trit with an associated confidence and weight—into a single result with
quorum, dissent, and abstain statistics:

\begin{equation}
  \hat{t} = \mathrm{sign}\!\left(
    \frac{\sum_i w_i \cdot c_i \cdot t_i}{\sum_i w_i \cdot c_i}
  \right)
  \label{eq:coalition}
\end{equation}

where $t_i \in \{-1,0,+1\}$, $c_i \in [0,1]$ is confidence, and
$w_i > 0$ is importance weight.

\subsection{Action Gate}

The action gate enforces a structural separation between hard constraints
and soft preferences. Each \texttt{GateDimension} carries an evidence
signal, a weight, and an optional \texttt{hard\_block} flag. A hard-block
dimension with negative evidence vetoes the action unconditionally,
regardless of all other dimensions:

\begin{equation}
  \text{verdict} = \begin{cases}
    \text{Block} & \exists\, d_i : \text{hard\_block}(d_i) \wedge t_{d_i} = -1 \\
    \text{sign}(\bar{S}) & \text{otherwise}
  \end{cases}
  \label{eq:gate}
\end{equation}

This is the structural analogue of a safety interlock: the veto is
unconditional, not probabilistic.

\subsection{Scalar Temperature and Hallucination Score}

\texttt{scalar\_temperature()} bridges the ternary decision to the LLM
sampling temperature domain: affirm at high confidence maps to a low
temperature (focused generation), hold maps to a mid temperature
(exploratory generation), and reject maps to near-zero (cautious refusal).

\texttt{hallucination\_score()} maps the variance of an evidence signal
vector to a trust trit: low variance (consistent signals) yields trust $+1$;
high variance yields $-1$, signalling that the agent's signals are
incoherent and should not be acted upon.

% ─────────────────────────────────────────────────────────────────────────────
\section{MoE-13 Ternary Orchestrator}
\label{sec:moe}

The MoE-13 architecture~\cite{moe13_2026} is implemented in the
\texttt{ternlang-moe} crate. It realises a ternary Mixture-of-Experts head
whose routing, synthesis, safety gate, and memory are all expressed natively
in balanced ternary semantics.

\subsection{Six-Dimensional Competence Space}

Each expert is characterised by a \emph{competence vector}
$\mathbf{v} \in [-1,1]^6$ across six axes:

\begin{equation}
  \mathbf{v} = \bigl[
    v_\text{syntax},\,
    v_\text{world},\,
    v_\text{reason},\,
    v_\text{tool},\,
    v_\text{persona},\,
    v_\text{safety}
  \bigr]
  \label{eq:competence}
\end{equation}

The synergy between two experts is defined as the complement of their cosine
similarity—orthogonal experts are maximally complementary:

\begin{equation}
  \sigma(\mathbf{v}_i, \mathbf{v}_j)
    = \frac{1 - \cos(\mathbf{v}_i, \mathbf{v}_j)}{2}
    \;\in\; [0, 1]
  \label{eq:synergy}
\end{equation}

\subsection{Dual-Key Synergistic Routing}

For each candidate expert pair $(i, j)$, the routing score combines
relevance to the query vector $\mathbf{q}$ with inter-expert synergy:

\begin{equation}
  \mathrm{score}(i, j)
    = \rho_i(\mathbf{q}) \cdot \rho_j(\mathbf{q}) \cdot \sigma(\mathbf{v}_i, \mathbf{v}_j)
  \label{eq:routing}
\end{equation}

where $\rho_k(\mathbf{q}) = \max(0,\, \cos(\mathbf{v}_k, \mathbf{q}))$
is the relevance of expert~$k$ to query $\mathbf{q}$. The pair with the
highest score is selected. Crucially, two highly relevant but similar
experts are penalised by low $\sigma$; the router favours complementary
coverage over redundant strength.

\subsection{$1{+}1{=}3$ Triad Synthesis}

After expert evaluation, an emergent \emph{triad field} is synthesised from
the two selected competence vectors and the routing synergy:

\begin{equation}
  \mathbf{E}_k = \sigma_{ij} \cdot \frac{\mathbf{v}_i + \mathbf{v}_j}{2}
  \label{eq:triad}
\end{equation}

This is the formal expression of $1{+}1{=}3$: two expert signals at high
synergy produce a third signal—the emergent field $\mathbf{E}_k$—that
neither expert could produce in isolation. The field modulates the
subsequent vote and the LLM temperature hint.

\subsection{Safety Hard Gate and Axis-6 Veto}

Dimension~6 of every competence vector is the \emph{safety axis}. Before
any vote is computed, the safety projection of the triad field is evaluated:

\begin{equation}
  \text{veto} \iff E_{k,\text{safety}} < \theta_s
  \label{eq:veto}
\end{equation}

where $\theta_s$ is the configurable safety threshold (default $-0.3$).
A veto immediately returns trit $-1$ at confidence $1.0$ and logs the event
to the Axis-tier memory for audit. No downstream reasoning can override it.

\subsection{Hold State and Tiebreaker}

When the weighted vote yields trit $0$ or aggregate confidence falls below
a hold threshold, the orchestrator does not immediately return. It invokes a
tiebreaker expert—selected by highest reasoning-axis score among inactive
experts—and re-votes with up to $\max\_\text{active}=4$ experts total. Only
if the tiebreaker also fails to resolve does the orchestrator return the
hold state to the caller, signalling that further evidence is needed.

\subsection{Three-Tier Memory Mesh}

The orchestrator maintains three memory tiers:

\begin{itemize}
  \item \textbf{Node} (TTL: seconds): LRU-bounded volatile store for
        within-session context. Evicts on capacity overflow.
  \item \textbf{Cluster} (TTL: minutes): routing frequency counters enabling
        mode-collapse detection ($\text{risk} = \max\_\text{pair} / \text{total}$).
  \item \textbf{Axis} (persistent): immutable audit log of safety vetoes with
        timestamp, expert identity, and query hash. Global priors over expert
        performance.
\end{itemize}

\subsection{Nine-Step Inference Pipeline}

The full orchestration pass executes in nine deterministic steps:
(1)~encode query to evidence vector;
(2)~dual-key route to best expert pair;
(3)~evaluate both experts independently;
(4)~synthesise triad field via~(\ref{eq:triad});
(5)~safety hard gate via~(\ref{eq:veto});
(6)~weighted trit vote with synergy amplification;
(7)~hold detection;
(8)~optional tiebreaker invocation;
(9)~return \texttt{OrchestrationResult} and update all three memory tiers.

\subsection{MCP Exposure}

The orchestrator is exposed as three MCP tool calls:
\texttt{moe\_orchestrate} runs a full pass and returns the complete result
including routing pair, triad field, verdicts, temperature, and prompt hint;
\texttt{moe\_deliberate} wraps the EMA deliberation engine for multi-round
iterative reasoning; \texttt{trit\_action\_gate} exposes the hard-block gate
as a standalone safety check. Any MCP-compatible AI client connected to
\texttt{ternlang-mcp} gains access to the full ternary reasoning stack as
tool calls.

% ─────────────────────────────────────────────────────────────────────────────
\section{Related Work}
\label{sec:related}

\textbf{Balanced ternary computing.} Knuth~\cite{knuth1997art} provides the
mathematical foundation. The Setun computer (1958)~\cite{brousentsov2002setun}
demonstrated physical ternary hardware. Modern revivals include academic
hardware studies at the University of South-Eastern Norway~\cite{bos2020ternary},
which targets C-to-ternary compilation for EDA tools—an approach that inherits
binary-native semantics and misses the symmetric properties of balanced ternary
that ternlang is designed to exploit natively.

\textbf{Ternary emulators.} Several open-source projects implement ternary
VMs in Python or JavaScript (Brandon Smith's 9-trit RISC simulator; Owlet,
an S-expression ternary interpreter in Node.js). These provide single-layer
implementations without compiler, ML kernels, or hardware backends.
\texttt{ternlang-compat} provides assembler-level bridges to both.

\textbf{Ternary quantization for neural networks.} BitNet~\cite{ma2024era}
and subsequent work (e.g., BitNet b1.58~\cite{ma2024bitnet158}) demonstrate
that large language models can be trained with ternary weights while retaining
competitive perplexity. The present work is the first to surface this property
as a first-class ISA opcode (\texttt{TSPARSE\_MATMUL}) rather than a software
library optimization.

\textbf{Quantum ternary.} Qutrits (3-level quantum systems) map naturally to
trit values $\{|{-1}\rangle, |0\rangle, |{+1}\rangle\}$. The BET encoding and
\texttt{trittensor} type system are structurally compatible with qutrit state
spaces; we leave the formal mapping to future work.

% ─────────────────────────────────────────────────────────────────────────────
\section{The Ternary Computing Ecosystem}
\label{sec:ecosystem}

A stated goal of ternlang is to serve as a \emph{convergence point} for the
fragmented ternary computing field. Table~\ref{tab:ecosystem} maps active
projects to their interoperability path via \texttt{ternlang-compat}.

\begin{table}[h]
\centering
\small
\caption{Ecosystem compatibility map}
\label{tab:ecosystem}
\begin{tabular}{p{2.2cm}p{2.2cm}p{2.8cm}}
\toprule
Project & Interface & Bridge \\
\midrule
Brandon Smith 9-trit sim & \texttt{.tasm} assembly & \texttt{TasmAssembler} $\to$ BET bytecode \\
Owlet (S-expr) & S-expression syntax & \texttt{OwletParser} $\to$ ternlang AST \\
USN / EDA tools & C-to-ternary & Academic whitepaper; ISA interop \\
Trit-Rust crate & Rust i8 trits & Superseded by ternlang-core \\
Q-Ternary & Qutrits & \texttt{trittensor} state model (future) \\
\bottomrule
\end{tabular}
\end{table}

Beyond runtime interoperability, ternlang aims to serve as a
\emph{conceptual archive} for prior ternary computing work.
Brandon Smith's Python 9-trit RISC simulator demonstrated a complete
fetch-decode-execute cycle in balanced ternary—an existence proof that
received little downstream adoption due to the absence of a compiler
ecosystem. The Owlet interpreter proved that S-expression evaluation maps
cleanly to a three-valued substrate. Both contributions are honoured in
the \texttt{examples/} corpus: \texttt{09\_risc\_fetch\_decode.tern}
translates Smith's pipeline decision logic into native BET language syntax,
and \texttt{13\_owlet\_bridge.tern} models cooperative ternary evaluation
with hold-as-suspension semantics. In this way, prior work that was
``slowly forgotten'' acquires a working executable form—runnable on the
same VM that drives the MoE-13 orchestrator.

% ─────────────────────────────────────────────────────────────────────────────
\section{Implementation Status}
\label{sec:status}

Ternlang is implemented in Rust and comprises the following crates:

\begin{table}[h]
\centering
\small
\caption{Workspace crates and test counts}
\label{tab:status}
\begin{tabular}{p{2.4cm}lp{4.0cm}}
\toprule
Crate & Tests & Description \\
\midrule
\texttt{ternlang-core}    & 31  & Lexer, parser, AST, semantic checker, BET VM (51 opcodes) \\
\texttt{ternlang-ml}      & 21  & BitNet quantization, sparse/CSC matmul, deliberation engine, coalition vote, action gate \\
\texttt{ternlang-moe}     & 16  & MoE-13 orchestrator: dual-key routing, triad synthesis, 3-tier memory, 13 experts \\
\texttt{ternlang-hdl}     & 21  & Verilog-2001 codegen, RTL simulator \\
\texttt{ternlang-runtime} & 4   & Distributed TCP actor runtime \\
\texttt{ternlang-compat}  & 29  & \texttt{.tasm} assembler, Owlet S-expr parser \\
\texttt{ternpkg}          & 5   & Package manager, GitHub-backed registry \\
\texttt{ternlang-lsp}     & ---  & LSP 3.17 (hover, completion, diagnostics) \\
\texttt{ternlang-mcp}     & ---  & MCP server, 10 tools including \texttt{moe\_orchestrate} \\
\texttt{ternlang-api}     & ---  & REST API, multi-tenant key management \\
\bottomrule
\end{tabular}
\end{table}

The full test suite comprises \textbf{146+ tests}, all passing.
All nine crates are published to \texttt{crates.io} under open-core
licensing (LGPL-3.0-or-later for the compiler and CLI; BSL-1.1 converting
to LGPL in 2030 for the ML engine and MoE orchestrator). A companion
\texttt{examples/} directory contains twenty executable \texttt{.tern}
programs spanning aerospace, medicine, distributed systems, hardware
pipelines (CPU fetch-decode tribute to Brandon Smith), civic governance,
finance, and AI agent design—each with an \texttt{INDEX.md} cataloguing
patterns, attribution, and quick-reference tables. The repository is
publicly available at the URL in the author affiliation.

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion and Future Work}
\label{sec:conclusion}

We have presented Ternlang, a full-stack balanced ternary execution
architecture spanning language, ISA, virtual machine, hardware backend,
distributed runtime, and AI reasoning primitives. Three layers of
contribution are unified under the same foundational primitive—the trit
$t \in \{-1, 0, +1\}$ with an active neutral state:

\begin{enumerate}
  \item \textbf{Execution substrate}: \texttt{TSPARSE\_MATMUL} achieves a
        $2.27\times$ reduction in multiply operations at baseline sparsity;
        a CSC kernel with Rayon reaches $122\times$ at high sparsity.
  \item \textbf{Reasoning primitives}: the five-tool Ternary AI Reasoning
        Toolkit makes AI agent decision loops structurally three-valued
        rather than binary with a ternary overlay.
  \item \textbf{MoE orchestration}: the MoE-13 orchestrator~\cite{moe13_2026}
        selects complementary expert pairs via dual-key synergistic routing,
        synthesises an emergent triad field ($1{+}1{=}3$), enforces a
        hard safety veto, and maintains a three-tier memory mesh—
        all exposed as MCP tool calls to any AI client.
\end{enumerate}

Two near-term publication targets are pending. First, the VS Code extension
(\texttt{ternlang-0.1.0.vsix}) will be submitted to the VS Code Marketplace
under publisher \texttt{rfi-irfos}, making the LSP, syntax highlighting, and
completion available to all VS Code users without manual installation. Second,
the MCP server will be submitted to the MCP registry, allowing any Claude
Desktop or compatible AI client to acquire ternary reasoning capabilities by
adding a single configuration entry.

Further research directions include:

\begin{itemize}
  \item \textbf{FPGA synthesis}: full \texttt{bet\_processor} targeting
        Xilinx Artix-7 and Lattice ECP5 via the pure-Rust RTL simulator
  \item \textbf{Memristor backend}: integration with physical ternary storage
        via the USN uMemristorToolbox~\cite{bos2020ternary}
  \item \textbf{End-to-end training}: native ternlang training loop with
        BitNet-style gradient quantization and ternary weight persistence
  \item \textbf{Qutrit bridge}: formal mapping of \texttt{trittensor} to
        qutrit state spaces for quantum-adjacent hardware
  \item \textbf{Community bootstrap}: structured outreach to the USN group
        (Bos \& Gundersen) for joint hardware-software co-design, and
        to the Bitcoin/ternary quantization community around BitNet b1.58
  \item \textbf{MoE-13 REST endpoints}: non-streaming POST interface for
        synchronous orchestration results alongside the existing SSE stream
\end{itemize}

The ternary computing field has been fragmented for decades. Ternlang is
designed to be the substrate where those fragments converge—from ISA to
orchestrator, from single trit to 13-expert MoE.

% ─────────────────────────────────────────────────────────────────────────────
\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
